---
title: "Using the data.generation function"
author: "Susumu Shikano"
date: "Februar 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the data.generation() function from github.


```{r}
source("https://raw.githubusercontent.com/SuShikano/StatTeaching/main/func/data_generation.R")


```

## Generating two different datasets

Let's generate 2 different datasets (n.sim=2) with the sample size of 500 (sample.size=500) and with the true regression model with 
only one independent variable (n.iv=1) whose mean is 3 (x.mu=3) and variance is 1 (x.Sigma=as.matrix(1,nrow=1)). The true regression model has the parameter value 1 for the intercept and 5 for the slope (para=c(1,5)). Its error has the uniform distribution (err.dist = "uniform"") with the range of 20 centered on 0 (err.disp = 20).

```{r}

samples01 <- data.generation(sample.size=100,
                                  n.sim=2,
                                  n.iv=1,
                             x.mu=3,
                             x.Sigma=as.matrix(1,nrow=1),
                             para=c(1,5),
                             err.dist = "uniform",
                             err.disp = 20)
```

## Describing the generated data

The first generated data:

```{r}
head(samples01$generated.data[[1]])
```


... and the second generated data:

```{r}
head(samples01$generated.data[[2]])
```

You can also plot y and x1 of both datasets:

```{r}
par(mfrow=c(1,2))

for (i.fig in 1:2){
  plot(samples01$generated.data[[i.fig]]$y ~ samples01$generated.data[[i.fig]]$X1,
       xlab="X1",ylab="y",
       main=paste("generated data",i.fig)
       )
}
```


## Regressing y on X1.

By using each of the datasets, we can regress y on X1. As one can see below, both results are not identical with the true model $\hat y = 1 + 5 \times X1$.

```{r}

for (i in 1:2){
  print(paste("Generated data",i))
  print(
    summary(
      lm(y ~ X1, data=samples01$generated.data[[i]])
    )
  )
}

```

## Checking the performance of the regression models using 1000 datasets

We generate 1000 datasets (n.sim=1000) based on the same set-up above.

```{r}

samples02 <- data.generation(sample.size=100,
                                  n.sim=1000,
                                  n.iv=1,
                             x.mu=3,
                             x.Sigma=as.matrix(1,nrow=1),
                             para=c(1,5),
                             err.dist = "uniform",
                             err.disp = 20)


```

Using each of the generated datasets, we regress y on X1 and collect the results:


```{r}

all.intercepts <- all.slopes <- matrix(NA,nrow=1000,ncol=4)
for(i.data in 1:1000){
  this.result <-     
     summary(lm(y ~ X1, 
                data=samples02$generated.data[[i.data]]))$coefficients
  all.intercepts[i.data,] <- this.result[1,]
  all.slopes[i.data,] <- this.result[2,]
} 


```

Let's check the distribution of the point estimates.

```{r}

par(mfrow=c(1,2))
for (i.fig in 1:2){
  if (i.fig==1) to.plot <- all.intercepts[,1]
  if (i.fig==2) to.plot <- all.slopes[,1]
  
  hist(to.plot,
       main=c("intercept","slope")[i.fig],
       xlab="Estimates",
       freq=F)
  legend("topright",
         paste("mean",round(mean(to.plot),2)),
         bty="n")
  
}


```

The means of the estimates are very close to the true value. If you use infinitely large number of datasets, the means are identical with the true value (no bias).
